---
title: "snprelate"
author: "Kin Lau"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    self_contained: yes
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
params:
    in_vcf: ""
    outdir: ""
---

```{r starttime}
# save start time for script
start_ptm <- proc.time()
start_ptm

outdir <- params$outdir
dir.create(outdir, recursive=TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.path=file.path(outdir, "individual_figures/"))
```

# Packages loaded

```{r loadlibs, echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE}
library(SNPRelate)
library(ggplot2)
library(dplyr)
```

# Convert VCF to GDS

We keep only the biallelic variants.

```{r readvcf}
vcf.fn <- params$in_vcf

# convert to GDS
gds_file <- file.path(outdir, 'all.gds')

snpgdsVCF2GDS(vcf.fn, gds_file, method="biallelic.only")

```

# Open the GDS file

```{r opengds}
genofile <- snpgdsOpen(gds_file)

```

# LD-pruned

We use an LD threshold of 0.5. SNPrelate scans the genome with a sliding window, discarding SNPs within each window of the LD exceeds 0.5. 

We also keep only SNPs with a missing rate of 0 (no missing data).

```{r ldprune}
set.seed(1000)

# Try different LD thresholds for sensitivity analysis
snpset <- snpgdsLDpruning(genofile, ld.threshold=0.5, missing.rate=0) # using LD threshold default used in SNPhylo

# Get all selected snp id
snpset.id <- unlist(snpset)
```

## Identity-By-State (IBS)

Average IBS is a metric for each pair of samples describing the average porportion of shared alleles for each locus.

```{r ibs_mat}
set.seed(100)
ibs <- snpgdsIBS(genofile, num.thread=1, autosome.only=TRUE, missing.rate=0, snp.id = snpset.id)

# we find 1- ibs so that the values become 'distances' where higher values represent more dissimilar pairs of samples
one_minus_ibs <- 1 - ibs$ibs
colnames(one_minus_ibs) <- ibs$sample.id
rownames(one_minus_ibs) <- ibs$sample.id
```

### MDS

```{r mds}
loc <- cmdscale(one_minus_ibs, k = 2)
#x <- loc[, 1]; y <- loc[, 2]
df <- tibble::as_tibble(loc, rownames="sample.id") %>%
  dplyr::rename(Dimension1=V1, Dimension2=V2)
#df$sample.id <- ibs$sample.id

# plot(x, y, xlab = "", ylab = "",
#     main = "Multidimensional Scaling Analysis (IBS)")

ggplot(df, aes(x=Dimension1, y=Dimension2, label=sample.id)) + 
  geom_point() + ggrepel::geom_label_repel() + ggtitle("MDS (based on IBS)")

```

### Dendrogram

Make dendrogram using hierarchical clustering on the IBS.

```{r makedendro}
ibs.hc <- snpgdsHCluster(ibs)

rv <- snpgdsCutTree(ibs.hc)

```

```{r plot_dendro}
plot(rv$dendrogram, main="Dendrogram based on IBS")

```

## PCA

```{r pca}
#pca <- snpgdsPCA(genofile, snp.id=snpset.id, num.thread=1)
pca <- snpgdsPCA(genofile, num.thread=1, autosome.only=TRUE, missing.rate=0, snp.id = snpset.id)

# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))

tab <- data.frame(sample.id = pca$sample.id,
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    EV3 = pca$eigenvect[,3],
    EV4 = pca$eigenvect[,4],
    EV5 = pca$eigenvect[,5],
    stringsAsFactors = FALSE)
head(tab)

#plot(tab$EV1, tab$EV2, xlab="eigenvector 1", ylab="eigenvector 2")
ggplot(tab, aes(x=EV1, y=EV2, label=sample.id)) + geom_point() + ggrepel::geom_label_repel() + ggtitle("PC1 and 2")
ggplot(tab, aes(x=EV3, y=EV4, label=sample.id)) + geom_point() + ggrepel::geom_label_repel() + ggtitle("PC3 and 4")

```

# No LD pruning

Here we don't prune for LD but we keep the other filters of biallelic SNPs and no missing data.

## Identity-By-State (IBS)

Average IBS is a metric for each pair of samples describing the average porportion of shared alleles for each locus.

```{r ibs_mat_noLDpruning}
set.seed(100)
ibs <- snpgdsIBS(genofile, num.thread=1, autosome.only=TRUE, missing.rate=0)

# we find 1- ibs so that the values become 'distances' where higher values represent more dissimilar pairs of samples
one_minus_ibs <- 1 - ibs$ibs
colnames(one_minus_ibs) <- ibs$sample.id
rownames(one_minus_ibs) <- ibs$sample.id
```

### MDS

```{r mds_noLDpruning}
loc <- cmdscale(one_minus_ibs, k = 2)

df <- tibble::as_tibble(loc, rownames="sample.id") %>%
  dplyr::rename(Dimension1=V1, Dimension2=V2)

ggplot(df, aes(x=Dimension1, y=Dimension2, label=sample.id)) + 
  geom_point() + ggrepel::geom_label_repel() + ggtitle("MDS (based on IBS)")

```

### Dendrogram

Make dendrogram using hierarchical clustering on the IBS.

```{r makedendro_noLDpruning}
ibs.hc <- snpgdsHCluster(ibs)

rv <- snpgdsCutTree(ibs.hc)

```

```{r plot_dendro_noLDpruning}
plot(rv$dendrogram, main="Dendrogram based on IBS")

```

## PCA

```{r pca_noLDpruning}
#pca <- snpgdsPCA(genofile, snp.id=snpset.id, num.thread=1)
pca <- snpgdsPCA(genofile, num.thread=1, autosome.only=TRUE, missing.rate=0)

# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))

tab <- data.frame(sample.id = pca$sample.id,
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    EV3 = pca$eigenvect[,3],
    EV4 = pca$eigenvect[,4],
    EV5 = pca$eigenvect[,5],
    stringsAsFactors = FALSE)
head(tab)

#plot(tab$EV1, tab$EV2, xlab="eigenvector 1", ylab="eigenvector 2")
ggplot(tab, aes(x=EV1, y=EV2, label=sample.id)) + geom_point() + ggrepel::geom_label_repel() + ggtitle("PC1 and 2")
ggplot(tab, aes(x=EV3, y=EV4, label=sample.id)) + geom_point() + ggrepel::geom_label_repel() + ggtitle("PC3 and 4")

```


# Close GDS

```{r closegds}
closefn.gds(genofile)

```

# SessionInfo

```{r sessioninfo}
sessionInfo()
```

# Time

```{r endtime}
# output time taken to run script
end_ptm <- proc.time()
end_ptm
end_ptm - start_ptm

```
